---
title: "Exercise: a complete LDL analysis"
author: "Dominic Schmitz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exercise: a complete LDL analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# What you will learn

This exercise introduces Linear Discriminative Learning (LDL) from the
beginning. No previous experience with LDL or Julia is assumed. By the end, you
will be able to

- represent word forms as an orthographic cue matrix **C**;
- represent meanings as a semantic matrix **S**;
- learn comprehension (**C -> S**) and production (**S -> C**) mappings;
- compute, inspect, combine, and save LDL measures; and
- fit frequency-informed and incremental mappings.

The code uses R throughout. `ldlr` sends computationally intensive operations
to Julia, JudiLing, and JudiLingMeasures in the background.

> **The two matrices are the heart of LDL.** Each row of **C** and **S** must
> describe the same lexical item, in the same order. **C** says which form cues
> are present; **S** says where that item's meaning lies in semantic space.

# Before you begin

This exercise assumes that you have

1. a data frame called `dat` with one observation per word or multiple
   observations containing a column called `word`;
2. a fastText text-vector file in `.vec` format; and
3. `ldlr` installed.

Replace the example file names and variable names where necessary.

```{r packages}
library(ldlr)

# Run this once after installing ldlr. It finds or installs Julia and installs
# the required Julia packages into ldlr's own environment.
setup_julia(install_julia = TRUE)

# Confirm that Julia was found and that the backend is ready.
ldlr_status()
```

If Julia is already installed, `setup_julia(install_julia = FALSE)` is enough.
An explicit executable path can be supplied with `setup_julia(julia = "...")`.

# 1. Minimal general workflow

## 1.1 Define the lexical items

LDL can be fitted at type level: one row represents one distinct word. Preserve
the order here, because this order will be used for both matrices.

```{r words}
words <- unique(as.character(dat$word))

length(words)
head(words)
anyDuplicated(words)       # should return 0
```

## 1.2 Build the cue matrix C

Here, word forms are represented by overlapping orthographic trigrams. Boundary
markers distinguish word edges. For example, `cat` yields `#ca`, `cat`, and
`at#`.

```{r cues}
cues <- make_ortho_trigrams(
  words,
  n = 3,
  boundary = "#",
  sparse = FALSE
)

C <- as.matrix(cues)

dim(C)
C[seq_len(min(5, nrow(C))), seq_len(min(10, ncol(C))), drop = FALSE]
```

`cues` is more than a matrix: it also stores the words, individual cue paths,
cue names, n-gram size, and boundary symbol. Keep this object. Cue-dependent
measures and path finding need that metadata later.

> **Task 1.** Inspect the first word in `words`. Identify its trigrams by hand,
> then locate their `1`s in the first row of `C`.

<details>
<summary><strong>One way to check your answer</strong></summary>

```{r task-1-solution}
words[1]
cues$paths[[1]]
names(which(C[1, ] == 1))
```

The last two results should contain the same cues.

</details>

## 1.3 Build the semantic matrix S

Each row of **S** is a semantic vector for one word. `load_fasttext_vectors()`
loads all vectors; indexing by `words` both selects the required vocabulary and
puts the rows in exactly the order used for **C**.

```{r semantics}
vectors <- load_fasttext_vectors("path/to/cc.en.300.vec")

# Check for absent words before subsetting. Otherwise missing names can be easy
# to overlook in a long analysis.
missing_words <- setdiff(words, rownames(vectors))
missing_words
stopifnot(length(missing_words) == 0L)

S <- vectors[words, , drop = FALSE]

dim(S)
stopifnot(
  nrow(C) == nrow(S),
  identical(rownames(C), rownames(S)),
  !anyNA(S),
  all(is.finite(S))
)
```

> **Task 2.** Explain why `vectors[rownames(vectors) %in% words, ]` is unsafe
> here even when it returns the correct number of rows.

<details>
<summary><strong>Explanation</strong></summary>

`%in%` retains the order of the complete fastText vocabulary, not the order of
`words`. Equal row counts therefore do not prove that rows refer to the same
items. Indexing with `vectors[words, ]` selects and orders rows simultaneously.

</details>

## 1.4 Learn comprehension and production

An end-state comprehension mapping learns a matrix **F** that transforms form
cues into predicted semantic vectors: **C F = S-hat**. An end-state production
mapping learns **G** in the reverse direction: **S G = C-hat**.

```{r endstate}
comprehension <- compute_comprehension(
  C = cues,
  S = S,
  learning = "endstate"
)

production <- compute_production(
  S = S,
  C = cues,
  learning = "endstate"
)

comprehension
production

summary(comprehension)
summary(production)
```

The printed `Accuracy` is an item-identification accuracy, not a regression
coefficient or percentage of variance explained. Because no test set was
supplied, it describes the same lexical items that were used to learn the
mapping and should not be interpreted as out-of-sample performance.

Useful components are available as ordinary R objects:

```{r inspect-models}
F <- coef(comprehension)
S_hat <- fitted(comprehension)

G <- coef(production)
C_hat <- fitted(production)

dim(F)
dim(S_hat)
dim(G)
dim(C_hat)
```

> **Task 3.** Use the dimensions to explain why **F** has one row per cue and
> one column per semantic dimension, whereas **G** has the reverse shape.

# 2. Measures

Measures turn the learned mappings and their predictions into item-level
quantities that can be inspected or used in later analyses. Some measures
compare predictions with targets; others describe the neighbourhood of an item
or the contribution of its cues.

```{r measure-catalog}
measure_catalog <- available_measures()
measure_catalog
```

The `stage` and `direction` columns show when each measure is available. For
example, a comprehension-only measure should not be requested from a production
model.

## 2.1 Compute all applicable measures

```{r compute-measures}
comprehension_measures <- compute_measures(
  comprehension,
  measures = "all",
  neighbours = 8,
  uncertainty_method = "correlation",
  functional_load_method = "correlation"
)

production_measures <- compute_measures(
  production,
  measures = "all",
  neighbours = 8,
  uncertainty_method = "correlation",
  functional_load_method = "correlation"
)

dim(comprehension_measures)
names(comprehension_measures)
head(comprehension_measures)

dim(production_measures)
names(production_measures)
head(production_measures)
```

## 2.2 Diagnose the results

Missing, infinite, or constant columns may be mathematically possible, but they
can also reveal incompatible data or a failed calculation. Always check them
before continuing.

```{r diagnose-measures}
check_measure_values(comprehension_measures)
check_measure_values(production_measures)

# A shorter display containing only questionable columns:
check_measure_values(comprehension_measures, problems_only = TRUE)
check_measure_values(production_measures, problems_only = TRUE)
```

Do not delete a problematic measure automatically. First determine why it is
missing or constant and whether that is meaningful for the data.

## 2.3 Save all measures in one data frame

Comprehension and production share some measure names. Prefixing them preserves
their direction and prevents accidental duplicate column names.

```{r combine-measures}
stopifnot(identical(
  as.character(comprehension_measures$item),
  as.character(production_measures$item)
))

comprehension_values <- comprehension_measures[, -1, drop = FALSE]
production_values <- production_measures[, -1, drop = FALSE]

names(comprehension_values) <- paste0("comprehension_", names(comprehension_values))
names(production_values) <- paste0("production_", names(production_values))

all_ldl_measures <- data.frame(
  item = comprehension_measures$item,
  comprehension_values,
  production_values,
  check.names = FALSE
)

dim(all_ldl_measures)
head(all_ldl_measures)

# Identify measures that contain more than one value per lexical item.
list_columns <- names(all_ldl_measures)[
  vapply(all_ldl_measures, is.list, logical(1L))
]
list_columns
```

Some cue-dependent measures are **list-columns**. For example, functional load
contains one value for every cue in a word, so different rows can contain
vectors of different lengths. This is valid inside an R data frame, but the CSV
format cannot represent such vectors as numeric values in a single cell.

Save the complete object as an RDS file. This preserves every scalar and every
cue-level vector exactly and can be read back with `readRDS()`.

```{r save-complete-measures}
saveRDS(all_ldl_measures, file = "all_ldl_measures.rds")

# Check that the saved object is identical to the original.
all_ldl_measures_reloaded <- readRDS("all_ldl_measures.rds")
stopifnot(identical(all_ldl_measures, all_ldl_measures_reloaded))
```

For a conventional CSV with exactly one scalar value per item and column,
exclude list-columns explicitly:

```{r save-scalar-measures}
scalar_ldl_measures <- all_ldl_measures[
  !vapply(all_ldl_measures, is.list, logical(1L))
]

write.csv(
  scalar_ldl_measures,
  file = "scalar_ldl_measures.csv",
  row.names = FALSE
)
```

If cue-level functional load is needed in CSV format, store it as a **long**
table: one row per item and cue. This preserves the numeric values instead of
turning a vector into text such as `c(0.32, 0.17, ...)`.

```{r save-functional-load}
functional_load_long <- do.call(
  rbind,
  lapply(seq_along(words), function(i) {
    values <- comprehension_measures$functional_load[[i]]
    data.frame(
      item = comprehension_measures$item[i],
      cue = cues$paths[[i]],
      cue_position = seq_along(values),
      functional_load = as.numeric(values),
      stringsAsFactors = FALSE
    )
  })
)

rownames(functional_load_long) <- NULL
head(functional_load_long)

write.csv(
  functional_load_long,
  file = "comprehension_functional_load.csv",
  row.names = FALSE
)
```

> **Task 4.** Open `scalar_ldl_measures.csv` and identify one comprehension
> measure and one production measure. Then inspect
> `comprehension_functional_load.csv`: how many rows belong to the first word,
> and how does that number relate to its trigrams? Why are both the direction
> prefixes and the long cue-level format important?

# 3. Path extension: from C-hat to a word form

The production mapping stops at **C-hat**, a graded vector of predicted cue
support. Producing an ordered word form requires an additional path-search step
that finds sequences of mutually compatible cues.

> **Warning: shown for reference only. Do not run this block during the
> exercise.** Path search can take many hours for a realistically sized
> vocabulary. Its duration depends on the vocabulary, cue inventory, algorithm,
> and search settings, so a short completion time cannot be guaranteed.

The following code is deliberately not evaluated when this page is built:

```{r path-extension, eval = FALSE}
forms <- produce_forms(
  production = production,
  comprehension = comprehension,
  cues = cues,
  method = "learn",
  max_candidates = 10,
  threshold = -1,
  verbose = TRUE
)

forms
head(forms$items)
head(forms$candidates)

form_measures <- compute_measures(forms, measures = "all")
check_measure_values(form_measures)
```

`forms$items` gives item-level results, while `forms$candidates` contains the
candidate paths retained by the search. Path-dependent measures become
available only after this extension.

# 4. Frequency-informed LDL

End-state learning treats every type once. Frequency-informed learning gives
more frequent types greater influence. The frequency vector must contain one
finite, non-negative value for every row of **C** and **S**, in their exact row
order.

If the original data contain multiple observations per word, first reduce the
frequency information to one value per type. The following example assumes that
`LogFreq.Zipf.` is constant within word type.

```{r frequencies}
dat$frequency <- round(exp(dat$LogFreq.Zipf.))

type_frequency <- aggregate(
  frequency ~ word,
  data = dat,
  FUN = function(x) x[1]
)

word_frequency <- type_frequency$frequency[
  match(words, type_frequency$word)
]

stopifnot(
  length(word_frequency) == length(words),
  !anyNA(word_frequency),
  all(is.finite(word_frequency)),
  all(word_frequency >= 0),
  identical(words, rownames(S)),
  identical(words, rownames(cues$C))
)
```

`match()` is crucial: it returns frequencies in `words` order rather than the
alphabetical order often produced by `aggregate()`.

```{r frequency-models}
comprehension_frequency <- compute_comprehension(
  C = cues,
  S = S,
  learning = "frequency",
  frequency = word_frequency
)

production_frequency <- compute_production(
  S = S,
  C = cues,
  learning = "frequency",
  frequency = word_frequency
)

comprehension_frequency
production_frequency
```

> **Task 5.** Compare the accuracy and selected measures of the end-state and
> frequency-informed models. Which differences do you observe? Why should a
> difference not automatically be called an improvement?

# 5. Incremental LDL

Incremental learning updates connection weights trial by trial. Therefore the
learning sequence, learning rate, number of epochs, and random seed matter. A
frequency vector can generate repeated exposures; using one exposure per word,
as below, provides a small first demonstration.

```{r incremental-models}
comprehension_incremental <- compute_comprehension(
  C = cues,
  S = S,
  learning = "incremental",
  frequency = rep(1L, length(words)),
  epochs = 2,
  learning_rate = 0.1,
  seed = 123
)

production_incremental <- compute_production(
  S = S,
  C = cues,
  learning = "incremental",
  frequency = rep(1L, length(words)),
  epochs = 2,
  learning_rate = 0.1,
  seed = 123
)

comprehension_incremental
production_incremental
```

The seed makes the randomized learning order reproducible. Keeping the seed,
epochs, and learning rate in an analysis script is therefore part of reporting
the model, not merely a technical detail.

> **Task 6.** Refit one incremental mapping with another seed. Then refit it
> with `seed = 123`. Which result is reproduced, and why?

# 6. Using LDL measures in regression analysis

_This section will be added later._

# What you should now be able to do

You should now be able to explain what **C**, **S**, **F**, **G**, **S-hat**, and
**C-hat** represent; fit comprehension and production mappings; compute and
diagnose measures; combine the measures in one clearly named data frame; and
distinguish end-state, frequency-informed, and incremental learning. The path
extension is a separate, potentially very expensive step and is not required
for analyses that stop at predicted cue support.
