---
title: "Learning methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Learning methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Both `compute_comprehension()` and `compute_production()` expose the same three
learning methods. The only difference is the direction of the mapping.

## End-state learning

```{r}
comp <- compute_comprehension(cues, S, learning = "endstate", shift = 0.02)
prod <- compute_production(S, cues, learning = "endstate", shift = 0.02)
```

End-state learning estimates all rows simultaneously. In R, `ldlr` validates
and aligns `C` and `S`, sends numeric matrices to Julia, and returns the mapping
and predictions as an R object. Julia calls
`JudiLing.make_transform_matrix(X, Y; shift = shift)`. The positive `shift`
regularizes the inverse; `0.02` is JudiLing's default and should generally be
kept constant when results are compared.

## Frequency-informed end-state learning

Frequency-informed learning gives each type a weight. The frequency vector must
have one non-negative value per matrix row and be in exactly the same order.

```{r}
type_frequency <- aggregate(
  frequency ~ word,
  data = data,
  FUN = function(x) x[1]
)
frequency <- type_frequency$frequency[match(rownames(S), type_frequency$word)]

stopifnot(length(frequency) == nrow(S), !anyNA(frequency))

comp_frequency <- compute_comprehension(
  cues, S, learning = "frequency", frequency = frequency
)
```

Do not exponentiate or round a logarithmic frequency variable unless that is
the theoretically intended transformation. In Julia this calls the weighted
method `JudiLing.make_transform_matrix(X, Y, frequency; shift = shift)`.

## Incremental learning

```{r}
comp_incremental <- compute_comprehension(
  cues, S,
  learning = "incremental",
  frequency = frequency,
  epochs = 2,
  learning_rate = 0.1,
  seed = 123
)
```

Incremental learning updates the weights event by event. If `frequency` is
provided, Julia calls `JudiLing.make_learn_seq()` to turn type frequencies into
a shuffled learning sequence using `seed`. It then calls
`JudiLing.wh_learn(X, Y; n_epochs, eta, learn_seq)`. Supply
`learning_sequence` explicitly to control event order; explicit order takes
precedence over a sequence generated from frequency.

```{r}
sequence <- c(1L, 2L, 1L, 3L, 4L)
comp_sequence <- compute_comprehension(
  cues, S,
  learning = "incremental",
  learning_sequence = sequence,
  epochs = 1,
  learning_rate = 0.1
)
```

The default sequence visits every row once when neither frequency nor an
explicit sequence is supplied. End-state and incremental estimates answer
different learning questions and should not be treated as interchangeable
numerical solvers.

## Train and test data

All learning methods accept a separately supplied evaluation set:

```{r}
cue_sets <- make_combined_ortho_trigrams(train_words, test_words)
S_train <- vectors[train_words, , drop = FALSE]
S_test  <- vectors[test_words, , drop = FALSE]

comp_test <- compute_comprehension(
  cue_sets$train, S_train,
  C_test = cue_sets$test, S_test = S_test,
  learning = "endstate"
)
```

`make_combined_ortho_trigrams()` is native R code and guarantees identical cue
columns across partitions. Fitting still uses the selected JudiLing operation;
R computes held-out predictions with `%*%` and records that the model was
evaluated on test data.

